---
title: "Week 2"
subtitle: "R basics"
---

# Basic operations

While much more versatile, R can be used as a calculator with some basic syntax.

- `+` - addition
- `-` - subtraction
- `*` - multiplication
- `/` - division
- `^` - raise to a power. E.g., `2^4` will compute $2^4 = 16$

:::{.callout-note title="Your Turn" icon=false}

What do the `%%` and `%/%` operators do? Try experimenting a bit. For example, what's the result of

```r
7 %% 3
```

and

```r
7 %/% 3
```
? Repeat with some different values to see if you can figure it out.
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
- `%%` - Gives the remainder of a division operation (i.e., "x modulo y").
- `%/%` - Gives the integer division of the left hand-side over the right-hand side (i.e., the "floored" result of $x/y$.)
:::

## Variable assignment

One of the key aspects of coding in R is *assignment*. That is, assigning objects or values to a name (we will come back to objects shortly).

Any of `=`, `<-`, and `->` can act as assignment operators. For example, in

```{r}
a <- 100
b = 100
100 -> c
```

all of `a`, `b`, and `c` are assigned the value 100. We can then use these objects in other computations. 

:::{.callout-note title="Your Turn" icon=false}

Create a new object named `ans` that is equal to $\frac{a + b}{c}$. Now, print `a`, `b`, `c`, and `ans` by simply typing the names of the objects into the console and pressing `ENTER`. Did the values of any of the original variables (`a`, `b`, and `c`) change?

:::

:::{.callout-warning title="Answer" icon=false collapse=true}
None of the original values change.
Code:
```r
ans <- (a + b) / c
```
:::

# Basic data types

There are a handful of basic data types in R, but the ones you will encounter most ofter are:

- integer
- numeric
- character
- logical

:::{.callout-note title="Your Turn" icon=false}
To find out the type or *class* of an object, you can use `class(object)`.
- What class is assigned to the special objects (called *constants*) `TRUE` and `FALSE`? 
- Run the code `z <- 2 + 3i`. What class is assigned to `z`?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
- The constants `TRUE` and `FALSE` get the class `logical`.
- `z` is assigned the class `complex`.
:::

## Classes can get "elevated"

Classes may change in R depending on what operation an object or value is involved in. Let's explore some scenarios in which this happens.

:::{.callout-note title="Your Turn" icon=false}
Create the following two objects in R:
```r
a <- 2L
b <- 2
```
1. What class is assigned to each of these objects?

Now, create a third object that is the sum of `a` and `b`.

```r
c <- a + b
```

2. What class is assigned to `c`? What happened?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
1. `a` has class `integer` and `b` has class `numeric`.
2. `c` has class `numeric`, meaning that `a` got elevated to `numeric` during this operation.
:::

:::{.callout-note title="Your Turn" icon=false}
Let's do another example, one that comes in handy for me all the time. What do you get when you type the following in the console and press enter?

```r
TRUE + FALSE
```

What about

```r
FALSE + FALSE + FALSE
```

or

```r
TRUE + TRUE
```
? How are the constants `TRUE` and `FALSE` being treated when involved in arithmetic operations?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
`TRUE` is given a value of 1 and `FALSE` is given a value of 0 when involved in arithmetic operations.
:::

# Objects and object classes

While there are some basic data types, there are many many classes of objects in R. To R, *everything* is an object, each with a specified class, and the class determines what sorts of operations an object can be involved in. Some other examples of object classes include:

- `data.frame`
- `matrix`
- `array`

:::{.callout-note title="Your Turn" icon=false}

In RStudio, type `is.` in the console and wait for the pop-up window of suggestions. Scroll through the possible classes.

What class is returned when you type `class(sum)`? What does this tell you?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
This tells us that functions are just another object class.
:::

## Methods for classes

Classes of objects have certain *methods* defined for them. These are generic functions, such as `summary()` or `print()`, that have specific instructions for how to handle different classes of objects. In order to see what functions work with a specific class of object, you can use the function `methods()`. This will give you an idea of what sorts of things you can do with a given object class.

:::{.callout-note title="Your Turn" icon=false}

Type `?methods()` into the console and press `ENTER` in order to learn how to use the function. How many methods are defined for the `data.frame` class? 

:::

:::{.callout-warning title="Answer" icon=false collapse=true}
There are 64 methods defined for the `data.frame` class.
:::

# Dataframes, vectors, and matrices

Some of the most common objects used in the R language are *dataframes*, *vectors*, and *matrices*. These are objects that can store *a lot* of data in a logical format. You have already seen how to read in data from a csv file in week 1. What we didn't cover was that the data were automatically stored in a *dataframe* when loaded into the environment.

## Working with vectors

Vectors are essentially lists of values. They can be lists of `character` values, `numeric` values, `logical` values, or even more general lists of just about any object type there is. These most general versions of vectors are called *lists* in R. 

```{r}
#| echo: false

# numeric vector
v <- c(1, 2, 3)

# logical vector
lv <- v < 2

# character vector
pets <- c("dog", "cat", "bunny")

# list is a special kind of vector
l <- list(v, lv, pets)
```


:::{.callout-note title="Your Turn" icon=false}

Run the following code.

```r
# numeric vector
(v <- c(1, 2, 3))

# logical vector
(lv <- v < 2)

# character vector
(pets <- c("dog", "cat", "bunny"))

# list is a special kind of vector
(l <- list(v, lv, pets))
```

🤔 How did we create the logical vector?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
The less-than sign, `<`, allowed us to ask whether each element of `v` was less than 2, resulting in a logical vector.
:::

Accessing elements of a vector is one of the most important things to know. Indexing in R is 1-based, meaning the first element of a vector is considered to be in position 1 (rather than position 0, like many other languages).

:::{.callout-note title="Your Turn" icon=false}

To access elements of a list or vector, we use square brackets, `[]`. Try the following:

```r
pets[1]
lv[3]
v[c(1,2)]
# or, the : character can make a range of values
v[1:3]
```

🤔 What happens if we make these negative? For example, `pets[-1]`?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
When the element number indicators are negative, R returns everything *except* that index.
:::

We can also combine multiple vectors into one. For example, try

```{r}
w <- c(4, 5, 6)
vw <- c(v, w)
vw
```

## Working with dataframes and matrices

Matrices and dataframes are 2-Dimensional generalizations of vectors, as if stacking vectors on top of each other or side by side. For example, two vectors,

$$
{\bf v} = \begin{bmatrix} 1 \\ 2 \\ 3 \end{bmatrix},\ \ {\bf w} = \begin{bmatrix} 4 \\ 5 \\ 6\end{bmatrix}
$$

can be concatenated into a single matrix

$$
{\bf M} = \begin{bmatrix}
1 & 4\\
2 & 5\\
3 & 6
\end{bmatrix}.
$$

In R, we can also have `character` and `logical` matrices.

:::{.callout-note title="Your Turn" icon=false}
Let's get familiar with these common R data structures. 

1. Create a dataframe in R from scratch.

```r
# combine vectors from above into a dataframe
df <- data.frame(
    v = v,
    w = w
)

# View the dataframe
View(df)

# convert to a matrix
M <- as.matrix(df)
M
```

Note how we can change the class of `df` and convert it into a matrix.

2. Let's create a new dataframe based on some different vectors.

```r
# add pets vector from above to a new dataframe
df2 <- data.frame(
    pet = pets,
    number = c(1, 0, 0)
)
df2

# now convert into a matrix
mat2 <- as.matrix(df2)
mat2
```

🤔 What happened when you converted the second dataframe into a matrix? What does this tell you about dataframes versus matrices?

> Hint: Use `class(df2$number)` and `class(mat2[, 2])`.
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
Upon converting a dataframe with numeric and character columns to a matrix, the numeric columns get elevated to a character so that all columns in the matrix are characters. This tells us that dataframes are more flexible than matrices and can store multiple data types in different columns.
:::

### Accessing elements, columns, and rows

Again being able to access, add, or remove elements (or even whole columns/rows) from data structures is one of the most important skills to learn. Try the following:

:::{.callout-note title="Your Turn" icon=false}
```r
mat2
mat2[1,2]
mat2[1, ]
mat2[, 2]
mat2[, "pet"]
mat2[, "number"]
```
🤔 Why do we need the comma when accessing components of a matrix? What goes before the comma and what goes after?

```r
df2
df2[1,2]
df2[1, ]
df2[, 2]
df2$number
df2$pet
```
As you can see, there are multiple ways to access different components of dataframes and matrices.

🤔 Does the `$` syntax work for matrices with named columns, like `mat2`?

:::

## Missing values

You will almost certainly have some missing data in your own graduate studies, either due to faulty instrumentation, a lapse in focus, or one of many other reasons. That is normal. So normal, that R uses a constant (a special object) `NA` to encode missing data. It's relatively easy to convert other codes for missing data (e.g., `999` or a blank cell in an excel spreadsheet) to `NA` values, once data are loaded into R or during the loading process. There are even methods designed to identify missing values. For example, try:

```{r}
is.na(c(2, 3, NA, 5, 6))
```

:::{.callout-note title="Your Turn" icon=false}
`NA` values can have unintended consequences depending on what functions you are using in your analysis code. For example, try the following:

```r
a <- c(2, 3, NA, 5, 6)
sum(a)
```

What do you get? What if you wanted to sum up all the values that are present and not missing? Type `?sum()`. What does the function documentation say to do to accomplish this?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
To get the total of the non-missing values, we can modify the code above to read `sum(a, na.rm = TRUE)`.
:::

:::{.callout-note}
 Note that the functions `mean()` and `sd()`, which take the mean and standard deviation of a vector of values, respectively, behave similarly.
:::

# Vectorization

R is optimized to work with vectors, matrices, and *arrays* (matrices but in even more dimensions). This is done through *vectorized* functions which apply the same function to all the elements of a vector or array at once. 

:::{.callout-note title="Your Turn" icon=false}

Let's use some of the vectors from above as examples. 

```r
# adding a constant to all the values in a vector
v + 10

# multiplying all elements by a constant
v * 2

# raising all elements to a power
v ^ 2

# many functions are also vectorized
sqrt(v)
```

🤔 What happens if you perform simple arithmetic operations with two vectors?

```r
# addition
v + w

# multiplication
v * w

# powers
v ^ w
```

🤔 What happens if you perform simple arithmetic operations with mixtures of vectors and matrices or dataframes?

```r
# vector plus a matrix?
v + mat

# multiplaction?
v * mat

# division?
v / mat
mat / v
```

🤔 What if the vector and matrix have differing numbers of rows?

```r
v2 <- c(v, 2)
v2 + mat
```
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
- When performing simple arithmetic operations on two vectors/matrices/dataframes with the same dimensions, R adds/divides/etc. corresponding elements.
- When performing simple arithmetic operations on a mixture of vectors and matrices where the length of the vector matches the number of rows in the matrix, R adds the vector to each column of the matrix, adding corresponding elements along the rows of each column.
- If the matrix and vector have differing numbers of rows, R starts to *recycle* the vector elements, starting with the upper left of the matrix as element 1, then moving down the rows before moving to the next column.
:::


# Working with "strings"

Learning how to work with strings is extremely valuable and can save you a lot of manual labor. We will see some more of how useful it can be next week, but first, let's cover some basics.

## Double versus single quotes

Most of the time, you can use double or single quotes interchangeably.

```{r}
str1 <- "This is a string."
str2 <- 'This is a string.'

# are they the same?
str1 == str2
```

*However*, if you want to include quotes inside the string, you need to use single quotes.

```{r}
str3 <- 'This is a "string".'
str3
```

Note that *printing* a string, as we did above with `str3`, is different from *writing* the string itself. To do that, we want

```{r}
writeLines(str3)
```

:::{.callout-note title="Your Turn" icon=false}

🤔 How did R internally modify our string `str3` above?

:::

:::{.callout-warning title="Answer" icon=false collapse=true}
R added *escape characters*, `\` to the string to *escape* the special characters.
:::

## Special characters and escapes

There are a handful of special characters, including things like `\`, `'`, and `"`. Other special characters you are likely to use are `\n` and `\t`, which are newline and tab characters, respectively. For example:

```{r}
str4 <- "Using\nnewline\ncharacters"
writeLines(str4)
```

To *escape* these special characters and use them literally in your string, precede them with `\`.

```{r}
writeLines("\'Escape\'")
writeLines("\"Escape\"")
writeLines("Some LaTeX: $\\bar{x} = \\frac{1}{n}\\sum_{i=1}^n x_i$")
```

We can also combine strings with other R output. For example:

```{r}
approx_pi <- round(pi, digits = 2)

writeLines(
    paste0("I ate some ", approx_pi, ".\n")
)

# using paste() with a separator between the strings
writeLines(
    paste("Lions", "Tigers", "and Bears", "oh my!", sep = ", ")
)
```

# Final thoughts

- Everything in R is an object with a specified class.
- Get used to working with dataframes and vectors. They will be your bread and butter for most analyses!



