---
title: "Intro2R Week 4"
subtitle: "Data visualization"
date: "8 October 2024"
format: md
toc: true
---


One of R's major strengths is in data visualization (i.e., plotting data). The most widely used data visualization package in the R community is `ggplot2`, which is part of `tidyverse`. Today's lab is all about data visualization.

# `ggplot2` basics

## The *grammar* of plotting

Hadley Wickham ([2009](https://doi.org/10.1198/jcgs.2009.07098)) is the author of `ggplot2` and designed the package to follow *The Grammar of Graphics* by Wilkinson (2005). 

>In brief, the grammar tells us that a graphic maps the data to the aesthetic attributes (colour, shape, size) of geometric objects (points, lines, bars).
> Wickham ([2015](https://ggplot2-book.org/)).

There is some vocabulary to go over.

(*Your notes below*)

- **Aesthetics**: 
- **Geometric elements** (or *geoms*):
- **Layers**:
- **Scales**:
- **Facets**:
- **Coordinate systems**:

## `ggplot2` syntax

- Data are mapped to aesthetics using the `aes()` function. Aesthetics that can be specified using `aes()` include:
    - Position arguments: `x`, `y`, `xmin`, `xmax`, etc,
    - Color arguments: `colour`, `fill`, `alpha`
    - Shape and size arguments: `shape`, `size`, `linetype`, `linewidth`
    - Grouping argument: `group`

- Geometric elements are added through the `geom_` family of functions. Examples include (but see a full list [here](https://ggplot2.tidyverse.org/reference/)):
    - `geom_point()`
    - `geom_line()`
    - `geom_boxplot()`

- Layers are added by placing as plus sign `+` between geometric elements.

- Facetting can be done using the `facet_wrap()` and `facet_grid()` functions.

- Scales can be changes using the family of `scale_` functions, including
    - `scale_y_continuous()`
    - `scale_color_manual()`

- Coordinate systems can be changed using the `coord_` family of functions, including
    - `coord_cartesian()`
    - `coord_polar()`
    - `coord_flip()`

- Finally, themes can be adjusted for the overall appearance of the graphic using `theme()` and its many (*many*) arguments.

# `ggplot2` practice (the best way to learn)

Let's load some data to play with. For this, we will use data that are available through the `lterdatasampler` package in R.

```{r}
if(!require("lterdatasampler")){
  install.packages("lterdatasampler")
}
# other required packages
library(tidyverse)

luq_streamchem <- lterdatasampler::luq_streamchem
```


## A simple scatter plot

We can start simple with these data and create a x-y scatterplot to investigate the relationship between variables of interest.

The first step is to establish a default ggplot object. This is done through the `ggplot()` function and letting it know where the data will be coming from.

```{r}
na_v_k <- ggplot(data = luq_streamchem)
```

Now, let's add a layer. Recall that each layer is made up of aesthetic mappings, mapping data to aesthetics, each of which has its own scale, a geometric element, and any statistical transformations that need to be performed on the data prior to the mapping. This information is supplied through the `geom_` family of functions, in combination with `aes()`.

```{r}
na_v_k <- na_v_k + 
  geom_point(aes(x = na, y = k))

na_v_k
```

Alternatively, we can combine these two steps into one with

```{r}
na_v_k <- ggplot(data = luq_streamchem, aes(na, k)) +
  geom_point()

na_v_k
```

In the above code block, we specify a global aesthetic mapping by putting the `aes()` call inside the `ggplot()` function rather than specific to a layer. This allows us to add additional layers, all using the same aesthetic mappings. For example,

```{r}
na_v_k + geom_point(colour = "white", size = 0.75)
```

::: {.callout-note}
Note how we can set fixed aesthetics inside the `geom_` function without needing to wrap them in `aes()`. If we want these aesthetics to encode information from the data, we need to map data to the aesthetics using `aes()`. We will see many examples of this soon.
:::

## Figures for distributions

Another useful type of plot is a histogram or density plot (a *smoothed* version of a histogram). Histograms are easy to plot in base R, but let's also see how to plot them in using `ggplot2()`. While `ggplot2()` requires more typing to create a *histogram*, it makes creating a *density plot* easier than with base R (in my opinion).

```{r}
ggplot(data = luq_streamchem, aes(x = na)) +
  geom_histogram()
```
:::{.callout-note}
Note how we don't need a `y` position aesthetic for a histogram, since that is actually implied by the *statistical transformation* component of the layer, which specifies that we want the count of the number of observations that fall within a given bin.
:::

We can create a density plot in a similar fashion.

```{r}
ggplot(data = luq_streamchem, aes(x = na)) +
  geom_density()
```

Let's customize this just a bit in order to get an introduction into how to change the aesthetic mappings to something you find pleasing to the eye.

```{r}
ggplot(data = luq_streamchem, aes(x = na)) +
  geom_density(fill = "brown", color = "darkgrey", alpha = 0.8)
```

:::{.callout-note icon=false title="Your Turn"}
Try changing the `fill`, `color`, and `alpha` aesthetics for a histogram as well.

🤔 What is the difference between `fill` and `color` for a these distribution plots? 
:::

:::{.callout-warning icon=false title="Answer" collapse=true}
`fill` changes the color of the middle of the geometry while `color` changes the outline of the geometry.
:::

## Figures comparing groups

Let's compare some groups of sugar maple seedlings in Hubbard Brook Experimental Forest, some growing in sites treated with supplemental calcium, others growing in reference sites. Sites were sampled in two years, 2003 and 2004.

```{r}
# first, load the data
data(hbr_maples) 

```


### Boxplots

Let's first use a box-plot to investigate seedling height as it varies by treatment group using a boxplot.
 
```{r}
ggplot(data = hbr_maples, aes(x = watershed, y = stem_length)) +
  geom_boxplot()
```

Here, we are comparing all seedlings from each watershed type from each year, but it might be interesting to see if there is year-to-year variation as well.

:::{.callout-info title="Your Turn" icon=false}
🤔 How might we "add" `year` to the plot such that we can compare the distribution of stem lengths across treatment types *and* year in the same plot?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
One option would be to map `year` to a *color* or *fill* aesthetic. Another might be to *facet* by `year`.
:::

```{r}
# the first thing we should do is check that `year` is treated
# as a categorical variable
hbr_maples %>% pull(year) %>% class()

# convert to a factor
hbr_maples <- hbr_maples %>% mutate(
  year_f = as.factor(year)
)
```

Above we converted `year` to a factor because we noticed it was loaded as numeric, but let's just see the type of warning message you will encounter if you forget to do something like this in the future.

```{r}
# example supplying numeric to a grouping factor
ggplot(data = hbr_maples, aes(watershed, stem_length, color = year)) +
  geom_boxplot()

# correctly supplying a factor
sl_box <- ggplot(data = hbr_maples, aes(watershed, stem_length, color = year_f)) +
  geom_boxplot()
sl_box
```

Let's add another layer to this plot to show where the actual data fall along the y-axis. This can be accomplished with `geom_jitter()`, which *jitters* points along the x-axis so that they don't overlap in one line up and down the y-axis.

```{r}
sl_box +
  geom_point(position = position_jitterdodge())
```

The above plot gets into adjusting the *position* aesthetic. Under the hood, `geom_boxplot()` used `position_dodge()` so that the boxplots for two years but the same treatment were not overlapping one another. We needed to match this position aesthetic when we added points to the plot, but we also wanted to add *jitter* to the position to so that the points are not all on top of one another.

### Density plots

We can also use density plot to compare distributions of the two treatment groups.

```{r}
ggplot(hbr_maples, aes(x = stem_length, fill = watershed)) +
  geom_density(alpha = 0.5)
```

:::{.callout-note title="Your Turn" icon=false}
🤔 How might we also add the `year` information to this plot?
:::

:::{.callout-warning title="Answer" icon=false collapse=true}
One option in this case would be to facet by either `year` or `watershed`.
:::

```{r}
ggplot(hbr_maples, aes(x = stem_length, fill = watershed)) +
  facet_wrap(vars(year), ncol = 2) +
  geom_density(alpha = 0.5)
```

### Plots with errorbars

A commonly used plot in the literature is called a *pointrange plot* in `ggplot2()`. For these plots, we want to plot a point estimate for each group (e.g., the mean) as well as error bars around that point. First, we need to do some data wrangling. Let's summarize the data by treatment and year groups and compute the mean and standard error of stem length for each group.

```{r}
# first get some summaries
hbr_sl_sum <- hbr_maples %>% 
  group_by(year_f, watershed) %>%
  summarise(
    mean = mean(stem_length),
    se = sd(stem_length) / sqrt(n()),
    .groups = "drop"
  )
hbr_sl_sum

# now create the plot
ggplot(hbr_sl_sum) +
  geom_pointrange(
    aes(
      x = watershed,
      y = mean,
      ymin = mean - se,
      ymax = mean + se,
      color = year_f
    ),
    position = position_dodge(width = 0.2)
  )

```


Alternatively, we could do a barplot with error bars. 

:::{.callout-note}
Note that `geom_bar()` counts the number of observations within a given category by default, while `geom_col()` allows you to map data to the height of a column.
:::

```{r}
# set global aesthetics
ggplot(hbr_sl_sum, aes(x = watershed, y = mean, fill = year_f)) +
  # add column layer
  geom_col(
    position = position_dodge(width = 0.5),
    width = 0.3
  ) +
  # add some error bars
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    position = position_dodge(width = 0.5),
    # this width adjusts the horizontal lines on the errorbars
    width = 0.1
  )
```

As an alternative, we could add a *statistical transformation* to a bar layer where we override the default statistic and change it to be the mean. We then add a summary statistic layer where we specify the geometry as `errorbar` and use a `ggplot2` function called `mean_se()` to compute the mean plus or minus one standard error.

```{r}
ggplot(hbr_maples, aes(x = watershed, y = stem_length, fill = year_f)) +
  # geom bar with mean as the fun argument
  geom_bar(
    stat = "summary", 
    fun = mean, 
    width = 0.3,
    position = position_dodge(width = 0.5)
  ) +
  # add a statistical summary layer
  stat_summary(
    geom = "errorbar",
    fun.data = mean_se,
    position = position_dodge(width = 0.5),
    width = 0.1
  )
```

## Line plots and time series

Let's first continue working with the maple seedling data and investigate the relationship between seedling height (`stem_length`) and leaf area and whether that differs by treatment type. 

```{r}
ggplot(
  hbr_maples, 
  aes(stem_length, corrected_leaf_area, color = watershed)
) +
  geom_point()
```


Perhaps we should also keep with our approach of looking at variation across years as well as treatments.

:::{.callout-note icon=false title="Your Turn"}
:thinking_face: Provide two aesthetics we could map year to in order to display another axis of variation on the above scatter plot.
:::

:::{.callout-warning icon=false title="Answer" collapse=true}
One option would be to use different shapes for the points from different years. Another option would be to facet by year.
:::


```{r}
ggplot(
  hbr_maples, 
  aes(stem_length, leaf_dry_mass, color = watershed)
) +
  geom_point(aes(shape = year_f))
```


We can apply transformations to data *inside* `ggplot()` calls. To illustrate this and to stabilize the mean-variance relationship between these two variables, let's log-transform the `leaf_dry_mass` measurements.

```{r}
ggplot(
  hbr_maples, 
  aes(stem_length, log(leaf_dry_mass), color = watershed)
) +
  geom_point(aes(shape = year_f))
```

Let's also see how to add *trend lines* and *smooths* to the figure. This can be accomplished with the `geom_smooth()` function. The default uses a [*loess smoother*](https://en.wikipedia.org/wiki/Local_regression), but a simple trend line looks appropriate to me here. To do that, we can set `method = "lm"`.

```{r}
ggplot(
  hbr_maples, 
  aes(stem_length, log(leaf_dry_mass), color = watershed)
) +
  geom_point(aes(shape = year_f)) +
  # add a layer for a smoother
  geom_smooth(aes(linetype = year_f), method = "lm")
```

:::{.callout-danger}
An important note here is that, while we get what look to be confidence bands with these figures, we should *almost never* use these in a publication graphic. These confidence bands come with a number of standard assumptions, but we have not checked them! It is always safer to provide your own confidence bands that come from rigorous model fitting and checking. This can be done with `geom_ribbon()`. So, let's remove those from the plot.
:::

```{r}
ggplot(
  hbr_maples, 
  aes(stem_length, log(leaf_dry_mass), color = watershed)
) +
  geom_point(aes(shape = year_f)) +
  # add a layer for a smoother
  geom_smooth(aes(linetype = year_f), method = "lm", se = FALSE)
```



```{r}
data("and_vertebrates")
```

As a final example, let's plot the number of cutthroat trout caught in different reaches of Mack Creek in the H. J. Andrews Experimental forest from 1987 to 2019.

```{r}
and_counts <- and_vertebrates %>%
  group_by(year, sitecode, species) %>%
  summarise(
    count = n(),
    .groups = "drop"
  ) %>% 
  drop_na()

```

```{r}
verts_ts <- filter(and_counts, species != "Cascade torrent salamander") %>%
ggplot(., aes(x = year, y = count, color = species)) +
  facet_wrap(vars(sitecode), ncol = 1) +
  geom_line() +
  geom_point()

verts_ts
```


## Customizing the appearance

To change the general appearance of the figures (background colors, grid lines, legend, etc.), we can change the *theme*. There are a handful of pre-made themes available, which you can view using `?ggplot2::theme_grey()`, which is the default. Out of these, I prefer `theme_classic()` or `theme_bw()` as a starting point. You can then adjust any of the them elements inside the `theme()` function.

```{r}
(verts_ts <- verts_ts + theme_bw())
```


That looks better than the grey background in my opinion, but what if we didn't want the gridlines in the background? This is something we can change with `theme()`. For a full list of elements you can modify with `theme()`, see the documentation with `?ggplot2::theme()`.

```{r}
# Remove minor gridlines
(verts_ts <- verts_ts +
  theme(
    panel.grid.minor = element_blank()
  ))
```

Let's also make the major vertical gridlines a bit darker.

```{r}
(verts_ts <- verts_ts + 
   theme(
     panel.grid.major.x = element_line(color = "darkgrey")
   ))
```


We can also adjust the background color of the facet labels with `theme()`.

```{r}
(verts_ts <- verts_ts + 
   theme(
     strip.background = element_rect(fill = "#D2AD7CFF")
   ))
```

### Adjusting scales

While aesthetics that are not related to data can be adjusted with `theme()`, if data are mapped to an aesthetic, we have to change the *scale* for that aesthetic in order to change its appearance. For example,

```{r}
(verts_ts <- verts_ts + 
   scale_color_manual(values = c("#BE7245FF", "#1C3333FF")))
```

Finally, the y-axis text is a bit scrunched, so let's change the breaks there using a `scale_` function.

```{r}
(verts_ts <- verts_ts + 
   scale_y_continuous(breaks = c(0, 100, 200)))
```

### Things to think about when customizing figures

- **Color blindness**: Are your colors color-blind friendly?

- **Business/ease of interpretation**: While figures with lots going on can be pretty to look at, is a busy figure the best way to convey the information you want readers to understand?

### Helpful resources

- **Geometry explorer**: [https://www.data-to-viz.com/](https://www.data-to-viz.com/)

- **Color palette packages**: [`paletteer`](https://pmassicotte.github.io/paletteer_gallery/)

- **Color palette explorer**: [https://r-graph-gallery.com/color-palette-finder](https://r-graph-gallery.com/color-palette-finder)

